refnx -- Neutron and X-ray reflectometry analysis in Python

### 2018-07-31 paper/manuscript.tex
PAPER: some replies to the polishing

\begin{abstract}
\refnx\ is a model-based neutron and X-ray reflectometry data analysis package written in Python. It is cross platform, and has been tested on Linux, macOS, and Windows. Its graphical user interface is browser-based, through a \Jupyter\ notebook.
Model construction is modular, being composed from a series of components that each describe a subset of the interface, parameterised in terms of physically relevant parameters (volume fraction of a polymer, lipid area per molecule, etc). The model and data are used to create an objective, which is used to calculate residuals, log-likelihood, and log-prior probabilities of the system. Objectives are combined to perform co-refinement of multiple datasets, and mixed-area models. Prior knowledge of parameter values is encoded as probability distribution functions or bounds on all parameters in the system. Additional prior probability terms can be defined for sets of components, over and above those available from the parameters alone. Algebraic parameter constraints are available.
A choice of fitting approaches is available, including least-squares (global and gradient-based optimizers) and a Bayesian approach using Markov Chain Monte Carlo to investigate the posterior distribution of the model parameters. The Bayesian approach is useful in examining parameter covariances, model selection, and variability in the resulting scattering length density profiles.
The package is designed to facilitate reproducible research; its use in \Jupyter\ notebooks, and subsequent distribution of those notebooks as supporting information, permits straightforward reproduction of analyses.
\end{abstract}

@@

The use of specular X-ray and neutron reflectometry for the morphological characterisation of thin films on the approximate size range 10 to \SI{5000}{\angstrom} has grown remarkably over the past years \cite{Wood2017, Daillant2009}. Most neutron and X-ray sources have instruments to perform reflectometry measurements, and there is an ongoing need for accessible software programs for users of those instruments to analyse their data in a straightforward fashion, including the co-refinement of multiple contrast datasets.  Several programs are available for this purpose, with a variety of different features \cite{Nelson2006,Bjorck2007,Kienzle2011,Gerelli2016}. These programs typically create a model of the interface, and either incrementally refine the model against the data using least-squares methods, or use Bayesian approaches \cite{Sivia2006, 2010arXiv1008.4686H} to examine the posterior probability distribution of the parameters (i.e.\ the statistical variation of the parameters in a model).

@@

\refnx\ is written in Python with an extensible object-oriented design, Figure~\ref{fig:components}, in which the user creates a model of the sample base on what they know about its composition and then hands that off to analysis code. As with \emph{Motofit} \cite{Nelson2006} it calculates reflectivity using the Abeles method \cite{Heavens1955} for specular reflection from a stratified medium.

@@

The building block of the analysis is the \Parameter\ object which represents a model value (e.g. the SLD of the material), whether it is allowed to vary in a fit, and a bounds attribute. The bounds are a probability distribution representing pre-existing knowledge of a parameter's value, called a prior probability.
A prior might be a simple uniform distribution that specifies a lower and upper bound (e.g. volume fraction is in the interval $[0, 1]$), or a normal distribution that represents an experimentally derived value and associated uncertainty (e.g. thickness is $\SI{100+-4}{\angstrom}$).
Any of the \emph{scipy.stats} \cite{Jones2001-2017} continuous distributions, or other distributions created by the user, can be used for this purpose. Algebraic relationships between \Parameter\ objects can be applied using constraints to permit more sophisticated constraints that cross between \Component\ objects to be described (e.g. the sum of the thicknesses of several layers is known to some uncertainty).

@@

The \Structure\ object represents the interfacial model, comprising of individual \Component\ objects in series. Each \Component\ represents a subset of the interface and selected attributes of the \Component\ can be described by physically relevant \Parameter\ objects. The simplest and most familiar \Component\ is a \Slab, which has a uniform scattering length density (SLD), thickness, roughness, and volume fraction of solvent. The simplest models are simply a series of \Slab\ objects.
More sophisticated components include \LipidLeaflet\ (a lipid monolayer) and \Spline\ (for freeform modelling of an SLD profile using spline interpolation).
It is straightforward to develop/modify new components for different structural functionality, a consequence of the program design.

To include further prior knowledge of the real sample into the model,
each \Component\ can additionally contribute to the prior probability in addition to its constituent \Parameter\ objects. This is useful when a \Component\ has a derived value, such as surface excess, which is already known, and can be included as prior knowledge of the system.

To calculate the reflectivity from series of \Component\ objects that form the model,
each \Component\ has a \emph{slabs} property that represents a discretised `slice' approximation to a continuous SLD profile for its particular region of the interface. A \Slab\ object has a single slice because it is a single thickness of uniform SLD. A \LipidLeaflet\ can be made of two slices (head/tail regions), but the \Spline\ has many thin slices approximating the smooth curve. Each of these slices has uniform SLD, with the N\'{e}vot--Croce approach being used to describe the roughness between them \cite{Nevot1980}.

@@

\subsection{Comparison with experimental data}

The \Objective\ class is the comparator of the predicted and measured reflectivities, using the \ReflectModel\ and a dataset, \DataD, to calculate $\chi^2$, log-likelihood (Equation~\ref{eqn:2}), log-prior, residuals, and the generative model.
The \DataD\ object has \emph{x, y, y\_err, x\_err} attributes to represent $Q$, $R$, $\mathrm{d}R$, $\mathrm{d}Q$. As is standard for many reflectometry data files, the \DataD\ object reads a three or four column plain-text datafile. A three column dataset represents $Q$ (\si{\per\angstrom}), $R$, $\mathrm{d}R$ (1 standard deviation). A four column dataset represents $Q$, $R$, $\mathrm{d}R$, $\mathrm{d}Q$ (\si{\per\angstrom}). Extending \DataD\ would allow other formats to be read. One example of this could be a wavelength dispersive file using ($\Omega$, $\lambda$)-data instead of $Q$, such as that used in energy scanned X-ray reflectometry, or sometimes produced by wavelength dispersive neutron reflectometers. In such a case \ReflectModel\ could be subclassed to make full use of this energy dispersive information.

An \Objective\ can be given a \Transform\ object to permit fitting as $\log_{10} R$ vs $Q$, $RQ^4$ vs $Q$; the default (no \Transform) is $R$ vs $Q$. Several \Objective\ objects can be combined to form a \GlobalObjective\ for co-refinement. The object-oriented nature allows \Parameter, \Component\ to be reused in different \Objective\ objects, and this is the basis for linking parameters between samples for co-refinement. For a comprehensive demonstration of multiple contrast co-refinement, see the annotated notebook in the supporting information.

\subsection{Statistical comparison and refinement}

@@

Confidence in the the results of the fitting come by examining the posterior probability, $p(\theta | I, D)$. In the simplest form, this is a confidence interval for the fitted parameter, but in the case of correlated parameters, a simple confidence interval can be rather misleading. The posterior probability is examined by encoding the likelihood and prior distributions and then using an MCMC algorithm (via the \emcee\ and \ptemcee\ packages) to perform affine invariant ensemble sampling \cite{emcee, ptemcee}
At the end of an MCMC run, the parameter set possesses a number of samples (called a `chain'), each of which are similarly plausible fits.
The statistics of the chain reveals the distribution and covariance of the parameters, the spread of the model-predicted measurements around the data, and in a reflectometry context, the range of structures that are consistent with the data.
The analysis of the chain thus yields a family of plausible SLD profiles and a family of plausible parameter ranges.
The chain statistics are used to update each \Parameter\ value, and assign a standard error. For the sampling, these represent the median and half the $[15.87, 84.13]$ percentile range respectively; the latter approximating the standard deviation for a normally distributed statistic.

The \ptemcee\ package is a variant (a `fork' in open-source software development terms) of the \emcee\ package that has been extended to implement the parallel tempering algorithm for characterisation of multi-modal probability distributions; different modes can be traversed by chain populations at higher `temperatures', while individual modes are efficiently explored by chains at lower `temperatures' \cite{ptemcee}. While \ptemcee\ is not strictly a global fitter, in practice, having multiple populations in the parallel tempering algorithm allows the fitter to converge to the global minimum and escape local minima.

Parallelisation of the sampling is automatic, making full use of multiprocessor machines and can use MPI on a cluster for yet greater parallelisation such as for corefinement of very large data sets. Visualisation of the samples produced by MCMC sampling is conveniently performed using the \corner\ package for scatter plot matrices \cite{corner}, which gives a representation of the probability distribution function for each individual parameter and also the correlation for each pair of parameters. As will be seen later, the plots for two normally distribute and uncorrelated parameters is isotropic, while correlated parameters show significant anisotropy. An evaluation of the impact of hard bounds can also be made by looking for correlation plots where the bounds are clearly truncating the distribution function, allowing the bounds to be re-evaluated and adjusted if necessary.

@@

A significant motivation in the development of \refnx\ has been the facilitation of reproducible analysis by helping the user describe \emph{how} the analysis was performed. A few lines of computer code is an incredibly powerful description, conveying the details with precision that is hard to match in written text and also being incredibly concise. Analysis code can be constructed in \refnx\ in relatively few lines of code (as illustrated in the supporting information), with example analyses often being sufficient to complete the task. These few lines of Python code can be further extended to produce publication quality plots saved and ready to import into the next publication.

While Python is a popular language for instruction and for data analysis, meaning that the relatively few lines of code required to complete a \refnx\ analysis of a set of experiments is not a huge hurdle, a simpler graphical user interface (GUI) is also provided.
The browser-based GUI is available for fitting within a \Jupyter\ notebook, Figure~\ref{fig:gui}, leveraging the ipywidgets modules \cite{ipywidgets}. The GUI has a `To code' button that turns the current model into the few lines of code required to perform the analysis without using the GUI, thus providing the desired instructions for the reproducible analysis. The ability to generate analysis code allows also makes it a stepping point for building more advanced models independently.
The current GUI is able to create slab based models for analysing a single dataset; a fully functional web-based reflectometry analysis notebook is currently available \cite{Nelson2018}.

@@

\begin{figure}
\centering
\label{fig:global_fit}%
\includegraphics[width=85mm]{./supporting_information/global_fit.pdf}%
%TODO: series labels on plots; delete points?;
%TODO: artefact in high q d2o fit; truncate at Q = 0/35 ?
%TODO: h2o fit isn't great about 0.20

@@

%TODO add (a) and (b)
\caption{a) Neutron reflectivity from a DMPC bilayer supported on a silicon crystal measured at three contrasts, with 500 samples from the posterior distribution in grey and median of the distribution in red. Data for the hdmix and \ce{H_2O} contrast offset by 0.1 and 0.001 respectively. b) SLD profile of the \ce{D_2O} model showing 500 samples from the posterior distribution, as well as the median in red. It is seen that the uncertainty in the reflectivity at high $Q$ is associated with an uncertainty in SLD profile at the lipid-\ce{D2O} interface.}
\end{figure}

\begin{figure}
  \includegraphics[width=175mm]{./supporting_information/corner.pdf}
  \caption{Corner plot for the varying parameters of DMPC bilayers measured at three contrasts. The sampling took $\sim$\SI{13}{\minute} on a \SI{3.2}{GHz} quad-core computer for 30 saved steps, corresponding to 6000 samples, with the steps being thinned by a factor of 100. A larger scale image is available in supporting information.}
  \label{fig:corner}
\end{figure}

>>>>

\begin{abstract}
\refnx\ is a model-based neutron and X-ray reflectometry data analysis package written in Python. It is cross platform, and has been tested on Linux, macOS, and Windows. Its graphical user interface is browser-based, through a \Jupyter\ notebook.
Model construction is modular, being composed from a series of components that each describe a subset of the interface, parameterised in terms of physically relevant parameters (volume fraction of a polymer, lipid area per molecule, etc). The model and data are used to create an objective, which is used to calculate residuals, log-likelihood, and log-prior probabilities of the system. Objectives are combined to perform co-refinement of multiple datasets, and mixed-area models. Prior knowledge of parameter values are encoded as probability distribution functions or bounds on all parameters in the system. Additional prior probability terms can be defined for sets of components, over and above those available from the parameters alone. Algebraic parameter constraints are available.
A choice of fitting approaches is available, including least-squares (global and gradient-based optimizers) and a Bayesian approach using Markov Chain Monte Carlo to investigate the posterior distribution of the model parameters. The Bayesian approach is useful in examining parameter covariances, model selection, and variability in the resulting scattering length density profiles.
The package is designed to facilitate reproducible research; its use in \Jupyter\ notebooks, and subsequent distribution of those notebooks as supporting information, permits straightforward reproduction of analyses.
\end{abstract}

@@

The use of specular X-ray and neutron reflectometry for the morphological characterisation of thin films on the approximate size range 10 to \SI{5000}{\angstrom} has grown remarkably over the past years \cite{Wood2017, Daillant2009}. Most neutron and X-ray sources have instruments to perform reflectometry measurements, and there is an ongoing need for accessible software programs for users of those instruments to analyse their data in a straightforward fashion, including the co-refinement of multiple contrast datasets.  Several programs are available for this purpose, with a variety of different features \cite{Nelson2006,Bjorck2007,Kienzle2011,Gerelli2016}. These programs typically create a model of the interface, and either incrementally refine the model against the data using least-squares methods, or use Bayesian approaches \cite{Sivia2006, 2010arXiv1008.4686H, Kienzle2011} to examine the posterior probability distribution of the parameters (i.e.\ the statistical variation of the parameters in a model).

@@

\refnx\ is written in Python with an extensible object-oriented design, Figure~\ref{fig:components}, in which the user creates a model of the sample based on what they know about its composition, with refinement of that model against the data. As with \emph{Motofit} \cite{Nelson2006} it calculates reflectivity using the Abeles method \cite{Heavens1955} for specular reflection from a stratified medium.

@@

The building block of the analysis is the \Parameter\ object which represents a model value (e.g. the SLD of the material), whether it is allowed to vary in a fit, and a bounds attribute. The bounds are a probability distribution representing pre-existing knowledge of a parameter's value, called a prior probability.
A prior might be a simple uniform distribution that specifies a lower and upper bound (e.g. volume fraction is in the interval $[0, 1]$), or a normal distribution that represents an experimentally derived value and associated uncertainty (e.g. thickness is $\SI{100+-4}{\angstrom}$).
Any of the \emph{scipy.stats} \cite{Jones2001-2017} continuous distributions, or other distributions created by the user, can be used for this purpose. Algebraic relationships between \Parameter\ objects can be applied to permit more sophisticated constraints that can cross between \Component\ objects (e.g. the sum of the thicknesses of several layers is known to some uncertainty).

@@

The \Structure\ object represents the interfacial model, assembled from individual \Component\ objects in series. Each \Component\ represents a subset of the interface and selected attributes of the \Component\ can be described by physically relevant \Parameter\ objects. The simplest and most familiar \Component\ is a \Slab, which has a uniform scattering length density (SLD), thickness, roughness, and volume fraction of solvent. The simplest models are simply a series of \Slab\ objects.
More sophisticated components include \LipidLeaflet\ (a lipid monolayer, or one-half of a lipid bilayer) and \Spline\ (for freeform modelling of an SLD profile using spline interpolation).
It is straightforward to develop/modify new components for different structural functionality, a consequence of the program design.

To include further prior knowledge of the real sample into the model,
each \Component\ can additionally contribute to the prior probability in addition to its constituent \Parameter\ objects. This is useful when a \Component\ has a derived value, such as surface excess, which is already known.

To calculate the reflectivity from series of \Component\ objects that form the model,
each \Component\ has a \emph{slabs} property that represents a discretised `slice' approximation to a continuous SLD profile for its particular region of the interface. A \Slab\ object has a single slice because it is a single thickness of uniform SLD. A \LipidLeaflet\ is made of two slices (head/tail regions), but the \Spline\ has many thin slices approximating the smooth curve. Each of these slices has uniform SLD, with the N\'{e}vot--Croce approach being used to describe the roughness between them \cite{Nevot1980}.

@@

\subsection{Model/data comparison}
The \Objective\ class is the comparator of the predicted and measured reflectivities, using the \ReflectModel\ and a dataset, \DataD, to calculate $\chi^2$, log-likelihood (Equation~\ref{eqn:2}), log-prior, residuals, and the generative model.
The \DataD\ object has \emph{x, y, y\_err, x\_err} attributes to represent $Q$, $R$, $\mathrm{d}R$, $\mathrm{d}Q$. As is standard for many reflectometry data files, the \DataD\ object reads a three or four column plain-text datafile. A three column dataset represents $Q$ (\si{\per\angstrom}), $R$, $\mathrm{d}R$ (1 standard deviation). A four column dataset represents $Q$, $R$, $\mathrm{d}R$, $\mathrm{d}Q$ (\si{\per\angstrom}). Extending \DataD\ would allow other formats to be read. One example of this could be a wavelength dispersive file using ($\Omega$, $\lambda$)-data instead of $Q$, such as that used in energy scanned X-ray reflectometry, or sometimes produced by wavelength dispersive neutron reflectometers. In such a case \ReflectModel\ could be subclassed to make full use of this energy dispersive information.

An \Objective\ can be given a \Transform\ object to permit fitting as $\log_{10} R$ vs $Q$, $RQ^4$ vs $Q$; the default (no \Transform) is $R$ vs $Q$. Several \Objective\ objects can be combined to form a \GlobalObjective\ for co-refinement. The object-oriented nature allows \Parameter, \Component\ to be reused in different \Objective, and this is the basis for linking parameters between samples for co-refinement. For a comprehensive demonstration of multiple contrast co-refinement, see the annotated notebook in the supporting information.

\subsection{Statistical comparison and model refinement}

@@

The posterior probability, $p(\theta | I, D)$, describes the distribution of parameter values consistent with the data and prior information. In the simplest form, this is akin to a confidence interval for a parameter derived by least-squares analysis. However, when parameters are correlated, or two models give similar quality of fit ('multi-modality'), a simple confidence interval can be misleading.
%FIXME - confidence intervals not used in Bayesian stats https://en.wikipedia.org/wiki/Credible_interval
 The posterior probability is derived by encoding the likelihood and prior distributions and then using an MCMC algorithm (via the \emcee\ and \ptemcee\ packages) to perform affine invariant ensemble sampling.\cite{emcee, ptemcee}
At the end of an MCMC run, the parameter set possesses a number of samples (called a `chain'); the samples reveal the distribution and covariance of the parameters, the spread of the model-predicted measurements around the data, and in a reflectometry context, the range of SLD profiles that are consistent with the data.
%FIXME the samples are not similarly plausible. 
The chain statistics are used to update each \Parameter\ value, and assign a standard error. For the sampling, these represent the median and half the $[15.87, 84.13]$ percentile range respectively; the latter approximating the standard deviation for a normally distributed statistic.

The \ptemcee\ package is a variant (a `fork' in open-source software development terms) of the \emcee\ package that has been extended to implement the parallel tempering algorithm for characterisation of multi-modal probability distributions; different modes can be traversed by chain populations at higher `temperatures', while individual modes are efficiently explored by chains at lower `temperatures' \cite{ptemcee}.  Having multiple populations in the parallel tempering algorithm allows the sampler to escape local maxima, greatly aiding it's ability to explore the most probable regions of the posterior.

Parallelisation of the sampling is automatic, making full use of multi-core machines, and can use MPI on a cluster for yet greater parallelisation. Visualisation of the samples produced by MCMC sampling is performed using the \corner\ package for scatter plot matrices \cite{corner}, which gives a representation of the probability distribution function for each individual parameter and also the covariance for each pair of parameters. As will be seen later, the plot for two normally distributed and uncorrelated parameters is isotropic, while covariant parameters show significant anisotropy. An evaluation of the impact of hard bounds can also be made by looking for plots where the bounds are clearly truncating the distribution function, allowing the bounds to be re-evaluated and adjusted if necessary.

@@

A significant motivation in the development of \refnx\ has been the facilitation of reproducible analysis by helping the user describe \emph{how} the analysis was performed. A few lines of computer code is an incredibly powerful description, conveying the details with precision that is hard to match in written text, as well as being incredibly concise.
%FIXME this line seems redundant.
%Analysis code can be constructed in \refnx\ in relatively few lines of code (as illustrated in the supporting information), with example analyses often being sufficient to complete the task.
These few lines of Python code can be further extended to produce publication quality plots saved and ready to import into the next publication.

While Python is a popular language for instruction and for data analysis, meaning that the relatively few lines of code required to complete a \refnx\ analysis of a set of experiments is not a huge hurdle, a simpler graphical user interface (GUI) is also provided.
The browser-based GUI is available for fitting within a \Jupyter\ notebook, Figure~\ref{fig:gui}, leveraging the \ipywidgets\ modules \cite{ipywidgets}. The GUI has a `To code' button that turns the current model into the few lines of code required to perform the analysis without using the GUI, thus providing the desired instructions for the reproducible analysis. The ability to generate analysis code allows also makes it a stepping point for building more advanced models independently.
The current GUI is able to use slab based models for fitting a single dataset; a fully functional web-based reflectometry analysis notebook is currently available \cite{Nelson2018}.

@@

\begin{figure}
\centering
\label{fig:global_fit}%
\includegraphics[width=85mm]{./supporting_information/global_fit.pdf}%
%TODO: series labels on plots; delete points?;
%FIXME: not clear what you mean by the above two points.

%FIXME: artefact in high q d2o fit; truncate at Q = 0/35 - I truncated the data at 0.3.
%FIXME: h2o fit isn't great about 0.20 - tried lots of things to improve this (including patchiness), nothing seems to work

@@

%TODO add (a) and (b)
%FIXME - where should I add (a) and (b), on the graph?
\caption{a) Neutron reflectivity from a DMPC bilayer supported on a silicon crystal, measured at three contrasts, with 500 samples from the posterior distribution in grey and median of the distribution in red. Data for the hdmix and \ce{H_2O} contrast offset by 0.1 and 0.001 respectively. b) SLD profile of the \ce{D_2O} model showing 500 samples from the posterior distribution, as well as the median in red. It is seen that the uncertainty in the reflectivity at high $Q$ is associated with an uncertainty in SLD profile at the lipid-\ce{D2O} interface.}
\end{figure}

\begin{figure}
  \includegraphics[width=175mm]{./supporting_information/corner.pdf}
  \caption{Corner plot for the varying parameters of DMPC bilayers supported on a silicon crystal, measured at three contrasts. The sampling took $\sim$\SI{13}{\minute} on a \SI{3.2}{GHz} quad-core computer for 30 saved steps, corresponding to 6000 samples, with the steps being thinned by a factor of 100. A larger scale image is available in supporting information.}
  \label{fig:corner}
\end{figure}
